Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-25 Manuscript/manuscript.Rmd
editted first 4 paragraphs of results; changed imbalance ratio to proportional imbalance

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. Here, the imbalance ratio is defined as the ratio of majority classes (regions not including TAD boundaries) to minority classes (regions including TAD boundaries) [@Orriols-Puig2008]. Such heavily imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models.

@@

We analyzed publically available replicate HiC data sets from two cell lines (Supplementary Table 1): GM12878 and K562. Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at four resolutions: 10 kb, 25 kb, 50 kb, and 100 kb. To account for the varying levels of data resolution, we developed a resolution-specific approach toward constructing TAD vs non TAD boundary regions on the linear genome. TAD boundary regions were defined as regions on the linear genome marking the coordinates of flanked TAD boundaries (Figure 1 A). To account for the varying levels of data resolution, we developed a resolution-specific approach toward constructing TAD vs non TAD boundary regions on the linear genome (Figure 1 B).

![**An illustration of the proposed approach for resolution-specific data construction used for TAD boundary prediction.** (A) Resolution-specific TADs were called and unique boundaries were identified. Boundaries were flanked on either side according to the Hi-C data resolution (i.e. 100 kb resolution corresponded to a 100 kb flank for a total boundary region of 200 kb). (B) The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a variable that indicated whether a genomic bin overlapped with a flanked TAD boundary ($Y = 1$) or not ($Y = 0$).](Figures/figure1_model_construction_V5.png)

## Number of TAD boundary regions decrease as Hi-C data resolution decrease

## TAD boundary regions exhibit cell-line specificity

## IMBALANCE RATIOS REMAINED SEVERE IRRESPECTIVELY OF Hi-C DATA RESOLUTION

To investigate class imbalance, defined as the ratio of TAD vs. non-TAD regions, we estimated the number of TAD boundaries ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) at each of the four resolutions of Hi-C data. Expectedly, the average number of TAD boundaries and the rest of genomic bins increased with the increased resolution of Hi-C data (Table 1) irrespectively of cell line (Supplementary Tables 3). However, imbalance ratios were found to be invariant to resolution, remaining consistently high with very little variability between chromosomes, regardless of cell line.

![**Cell-line specific boundary summaries.** (A) Barplots illustrating that the total number of TAD boundary regions decreases as Hi-C data resolution decreases. (B) Venn diagrams showing that TAD boundary regions exhibit cell-line specificity across resolution. (C) Boxplots indicating that imbalance ratios remain severe irrespective of resolution.](Figures/bounds_and_irs.png)

## IMBALANCE RATIOS REMAINED SEVERE IRRESPECTIVELY OF Hi-C DATA RESOLUTION

To investigate class imbalance, defined as the ratio of TAD vs. non-TAD regions, we estimated the number of TAD boundaries ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) at each of the four resolutions of Hi-C data. Expectedly, the average number of TAD boundaries and the rest of genomic bins increased with the increased resolution of Hi-C data (Table 1) irrespectively of cell line (Supplementary Tables 3). However, imbalance ratios were found to be invariant to resolution, remaining consistently high with very little variability between chromosomes, regardless of cell line.

>>>>

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. Specifically, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non boundary regions. Here, the proportion of imbalance is defined as the ratio of the set of TAD boundary regions to the sum of both TAD and non-TAD boundary regions.<!--Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. Here, the imbalance ratio is defined as the ratio of majority classes (regions not including TAD boundaries) to minority classes (regions including TAD boundaries) [@Orriols-Puig2008].--> Such heavily imbalanced datasets are known to adversely impact the performance of classifiers as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models.

@@

We analyzed publically available replicate Hi-C data sets from two cell lines including GM12878 and K562 (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at four resolutions: 10 kb, 25 kb, 50 kb, and 100 kb. To account for the varying levels of data resolution, we developed a resolution-specific approach toward constructing TAD versus non TAD boundary regions on the linear genome, labeled as a response vector **Y** (Figure 1). First we binned the linear genome according to the data resolution. We then flanked the boundaries on either side by the same number of basepairs as the previously constructed genomic bins. TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$).

![**An illustration of the proposed approach for resolution-specific data construction used for TAD boundary prediction.** Resolution-specific TADs were called and unique boundaries were identified, denoted by the blue dots. The yellow regions denote the flanked boundary regions according to the Hi-C data resolution (i.e. 100 kb resolution corresponded to a 100 kb flank for a total boundary region of 200 kb). The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a variable that indicated whether a genomic bin overlapped with a flanked TAD boundary ($Y = 1$) or not ($Y = 0$).](Figures/figure1_model_construction_V7.png)

## TOTAL NUMBER OF TAD BOUNDARY REGIONS DECREASE AS Hi-C DATA RESOLUTION DECREASES

Expectedly, the total number of TAD boundary regions increased with the increased resolution of Hi-C data irrespectively of cell line (Figure 2 A). At higher resolutions, GM12878 was found to contain more TADs, and thus more TAD boundary regions compared to K562. This can largely be attributed to the increased sequencing depth of the GM12878 library. However, this difference was found to dissipate as resolution decreased, resulting in much more similar numbers of TAD boundary regions between the two cell lines at the 50 kb and 100 kb resolutions respectively.

## TAD BOUNDARY REGIONS EXHIBIT CELL-LINE SPECIFICITY

We then sought to compare the cell-line specificity of TAD boundary regions. Between the two cell lines, a significant number of TAD boundary regions were specific to one of the two cell lines (Figure 2 B). It was found that 45%, 38%, 37%, and 36% of TAD boundary regions were specific to the GM12878 cell line, across each resolution respectively. Similarly, for K562, it was found that 27%, 27%, 29%, and 33% of TAD boundary regions were cell line specific for each resolution specifically.

## PROPORTIONAL IMBALANCE REMAINED SEVERE IRRESPECTIVELY OF Hi-C DATA RESOLUTION

We then investigated the proportion of class imbalance, defined as the ratio of TAD boundary regions versus the sum of both TAD and non-TAD boundary regions. We compared the estimated the number of TAD boundary regions ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) for each cell line, at each of the four resolutions of Hi-C data. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3 C).

![**Cell-line specific boundary summaries.** (A) Barplots illustrating that the total number of TAD boundary regions decreases as Hi-C data resolution decreases. (B) Venn diagrams showing that TAD boundary regions exhibit cell-line specificity across resolution. (C) Boxplots indicating that the proportion of class imbalance remained severe irrespective of resolution.](Figures/bounds_and_pis.png)

<!--
## IMBALANCE RATIOS REMAINED SEVERE IRRESPECTIVELY OF Hi-C DATA RESOLUTION

To investigate class imbalance, defined as the ratio of TAD vs. non-TAD regions, we estimated the number of TAD boundaries ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) at each of the four resolutions of Hi-C data. Expectedly, the average number of TAD boundaries and the rest of genomic bins increased with the increased resolution of Hi-C data (Table 1) irrespectively of cell line (Supplementary Tables 3). However, imbalance ratios were found to be invariant to resolution, remaining consistently high with very little variability between chromosomes, regardless of cell line.
-->
